Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-01-13 paper/paper.tex
More writing and drawing

\usepackage{showframe}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.
Despite these problems being \NP-complete, modern practical subgraph isomorphism algorithms can
handle problem instances with many hundreds of vertices in the pattern graph, and up to ten thousand
vertices in the target graph \citep{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}.

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets (cite ??), which were generated by
taking a random subgraph of a random graph (using various models, including scale-free and bounded
degree) and permuting the vertices.  This is not ideal: such instances are guaranteed to be
satisfiable (?? mention \citep{Anton:2009}), and so existing benchmark suites contain relatively few
non-trivial unsatisfiable instances. These instances also tend to be computationally fairly easy,
with most of the difficulty being in dealing with the size of the model.

@@

@@

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\overline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ iff
$v$ and $w$ are not adjacent in $G$. We write $G(v, p)$ for a ?? ER random graph with $v$ vertices,
and an edge between each distinct pair of vertices with independent probability $p$.

@@

The \emph{association graph encoding} of a subgraph isomorphism problem is constructed by creating a
new graph with a vertex for each pair $(p, t)$ of vertices from the pattern and target graphs
respectively. There is an edge between vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$
to $t_1$ and $p_2$ to $t_2$ simultaneously is permitted. For the non-induced problem, permitted
means either $p_1$ is not adjacent to $p_2$, or both $p_1$ and $p_2$, and $t_1$ and $t_2$, are
adjacent; for the induced problem, it means that $p_1$ is adjacent to $p_2$ if and only if $t_1$ is
adjacent to $t_2$. Either way, it can be seen that a clique (a subset of vertices, each of which is
adjacent to every other in the subset) of size equal to the order of the pattern graph exists in the
association graph if and only if the problem is satisfiable, and the vertices selected correspond to
the assignments making the solution \citep{Levi:1973}.

@@

@@

\begin{figure}[tb]
    \input{gen-graph-non-induced.tex}
    \caption{Behaviour of algorithms on the non-induced variant. Each point is the average of ten
        runs. For each plot, the x-axis is the pattern edge probability and the y-axis is the target
        edge probability, both from 0 to 1. Along the top row, we show the proportion of instances which are
        satisfiable; the white bands shows the phase transitions. The second row is our prediction
        of what the first row will look like. On the third row, we show the number of search nodes
        used by the Glasgow algorithm, on the fourth row, the number of search nodes used by the LAD
        algorithm, and the fifth, VF2; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the phase transition for the non-induced variant, for
patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis) and target
(y-axis) edge probabilities. Inside the orange region, at the bottom right of each plot, every
instance is unsatisfiable---here we are trying to find a dense pattern in a sparse target. In the
purple region, at the top left, every instance is satisfiable---we are looking for a sparse pattern
in a dense target (which is easy, since we only have to preserve adjacency, not non-adjacency). The
white band between the regions shows the location of the phase transition: here, roughly half the
instances are satisfiable.

@@

\subsection{Locating the Phase Transition, and Heuristics}

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence, and drawing inspiration
from the association graph encoding. Since we are trying to find an \emph{injective} mapping from a
pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are $t^{\underline{p}} = t \cdot (t -
1) \cdot \ldots \cdot (t - p + 1)$ possible assignments of target vertices to pattern vertices.  We expect the
pattern to have $d_p \cdot \binom{p}{2}$ edges, so we obtain the probability of all of these edges
being mapped to edges in the target by raising $d_t$ to this power, giving an expected number of
solutions of \[ \langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}}
\textnormal{,} \] which we plot in the second row of \cref{figure:non-induced}.

@@

@@

\begin{figure*}[p]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs.}\label{figure:induced}
\end{figure*}

@@

@@

Note that this is solver-independent.

@@

By repeating the argument for locating the non-induced phase transition and additionally considering
non-edges, we get an expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}}
    \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}}
\textnormal{.} \] We plot this
on the second row of \cref{figure:induced}---again, our prediction is accurate except for very
sparse or very dense patterns.

?? Graph showing the product of satisfiable with vertical flip, min (unsat) and max (sat) of search
nodes with vertical flip.

This suggests that we're not able to make use of adjacency and non-adjacency simultaneously.

@@

@@

@@

@@

Direct pseudo-boolean encoding: variables for each assignment, cardinality constraints, edge and
non-edge constraints. We use the Clasp solver. We see that hard instances remain hard, including
inside the central region, and that the easy satisfiable instances remain easy. (Although not
presented here, we also tried an equivalent SAT encoding with the Glucose solver, and saw very
similar performance.)

We also tried using a mixed integer encoding along similar lines. Gurobi was only able to solve some
of the trivial satisfiable instances, and was almost never able to prove unsatisfiability within the
time limit. We do not plot these results.

Clique encoding: association graph, a version of BBMC modified to solve the decision problem. The
hard instances in the central region remain hard. Additionally, some of the easy satisfiable
instances become hard. Possible weakness in variable ordering heuristics.

>>>>

% \usepackage{showframe}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.
Despite these problems being \NP-complete, modern practical subgraph isomorphism algorithms can
handle problem instances with many hundreds of vertices in the pattern graph, and up to ten thousand
vertices in the target graph \citep{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading
to successful application in areas such as computer
vision \citep{Damiand:2011,Solnon:2015}, biochemistry \citep{Giugno:2013}, and pattern recognition
\citep{Conte:2004}.

@@

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices.  This is not
ideal: such instances are guaranteed to be satisfiable (?? mention \citep{Anton:2009}), and so
existing benchmark suites contain relatively few non-trivial unsatisfiable instances. These
instances also tend to be computationally fairly easy, with most of the difficulty being in dealing
with the size of the model.

@@

?? Maybe introduce \cref{figure:phase-transition} here instead of at the start of the next section?

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\overline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ iff
$v$ and $w$ are not adjacent in $G$. We write $G(v, p)$ for an Erd\H{o}s-R\'enyi random graph with
$v$ vertices, and an edge between each distinct pair of vertices with independent probability $p$.

@@

The \emph{association graph encoding} of a subgraph isomorphism problem is constructed by creating a
new graph with a vertex for each pair $(p, t)$ of vertices from the pattern and target graphs
respectively. There is an edge between vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$
to $t_1$ and $p_2$ to $t_2$ simultaneously is permitted. For the non-induced problem, permitted
means either $p_1$ is not adjacent to $p_2$, or both $p_1$ and $p_2$, and $t_1$ and $t_2$, are
adjacent; for the induced problem, it means that $p_1$ is adjacent to $p_2$ if and only if $t_1$ is
adjacent to $t_2$. Either way, it can be seen that a clique (a subset of vertices, each of which is
adjacent to every other in the subset) of size equal to the order of the pattern graph exists in the
association graph if and only if the problem is satisfiable, and the vertices selected correspond to
the assignments making the solution \citep{Levi:1973}. We illustrate this in
\cref{figure:sip}. ?? Can we move this to section 4? I thought we'd need this for the sol formulae,
but it appears not.

@@

?? Timeouts

@@

\begin{figure}[tb]
    \input{gen-graph-non-induced.tex}
    \caption{Behaviour of algorithms on the non-induced variant. Each point is the average of ten
        runs. For each plot, the x-axis is the pattern edge probability and the y-axis is the target
        edge probability, both from 0 to 1. Along the top row, we show the proportion of instances which are
        satisfiable; the white bands shows the phase transitions, and the black lines are our
        predictions of where the phase transition will occur. On the second row, we show the number
    of search nodes used by the Glasgow algorithm, on the third row, the number of search nodes used
by the LAD algorithm, and the fourth, VF2; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the phase transition for the non-induced variant, for
patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis) and target
(y-axis) edge probabilities. Inside the orange region, at the bottom right of each plot, every
instance is unsatisfiable---here we are trying to find a dense pattern in a sparse target. In the
purple region, at the top left, every instance is satisfiable---we are looking for a sparse pattern
in a dense target (which is easy, since we only have to preserve adjacency, not non-adjacency). The
white band between the regions shows the location of the phase transition: here, roughly half the
instances are satisfiable. (We discuss the black line below.)

@@

\subsection{Locating the Phase Transition}

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence, and drawing inspiration
from the association graph encoding. Since we are trying to find an \emph{injective} mapping from a
pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are $t^{\underline{p}} = t \cdot (t -
1) \cdot \ldots \cdot (t - p + 1)$ possible assignments of target vertices to pattern vertices.  We expect the
pattern to have $d_p \cdot \binom{p}{2}$ edges, so we obtain the probability of all of these edges
being mapped to edges in the target by raising $d_t$ to this power, giving an expected number of
solutions of \[ \langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}}
\textnormal{.} \] This formula predicts a very sharp phase transition from $\langle Sol \rangle \ll
1$ to $\langle Sol \rangle \gg 1$. We plot where this occurs using black lines in the first row of
\cref{figure:non-induced}.

@@

?? Also because of half edges.

\subsection{Variable and Value Ordering Heuristics}

?? Note that the Glasgow heuristics are moving us towards the satisfiable region, or increasing the
expected number of solutions \citep{Gent:1996:EN}.

?? LAD does SDF with no tiebreaking, no value ordering heuristic.

?? In contrast, VF2 heuristics work by selecting vertices adjacent to vertices already selected.

?? I suppose we should mention kappa \citep{Gent:1996:Kappa}, etc, here. This whole thing is a bit
subtle: maximising E(N) or Kappa as a principle is encoding-independent, but whether this is done by
SDF, deg, etc, is encoding-dependent.

@@

\begin{figure*}[tb]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second, third and fourth rows show the number of search nodes used by the
    Glasgow, LAD and VF2 algorithms. The fifth row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs. The final row shows when the Glasgow algorithm performs better when given the
    complements of the pattern and target graphs as inputs.}\label{figure:induced}
\end{figure*}

@@

?? Explain the flip bound on row 5, and note that the difficult region is roughly near the phase
transitions in the original graphs.

?? This suggests that we're not able to make use of adjacency and non-adjacency simultaneously. Note
that this is solver-independent. In the following section we see whether this is an oddity of
subgraph isomorphism algorithms.

@@

@@

By repeating the argument for locating the non-induced phase transition and additionally considering
non-edges, we get an expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}}
    \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}}
\textnormal{.} \] We plot this
using black lines on \cref{figure:induced}---again, our prediction is accurate except for very
sparse or very dense patterns.

@@

?? Explain why we might guess degree heuristics just won't work for induced.

?? Explain bottom row of \cref{figure:induced}, and give way of calculating it.

@@

?? Is this heuristic better in general, on non-random instances?

@@

?? The odd stuff in the middle, which isn't near a phase transition, but is hard.

@@

Direct pseudo-boolean encoding: variables for each assignment, cardinality constraints, edge and
non-edge constraints. We use the Clasp solver. We see that hard instances remain hard, including
inside the central region, and that the easy satisfiable instances remain easy.

Although not presented here, we tried an equivalent SAT encoding with direct encoding of the
cardinality constraints using the Glucose solver, and saw very similar performance.  We also tried
using a mixed integer encoding along similar lines---Gurobi was only able to solve some of the
trivial satisfiable instances, and was almost never able to prove unsatisfiability within the time
limit. We do not plot these results.

Clique encoding: association graph, a version of BBMC \citep{SanSegundo:2011} modified to solve the
decision problem. The hard instances in the central region remain hard. Additionally, some of the
easy satisfiable instances become hard. Possible weakness in variable ordering heuristics.
